Carrier phase and amplitude estimation for phase shift keying using pilots and data

### 2013-11-19 response.tex

\renewcommand{\theenumii}{(\alph{enumii})}
\renewcommand{\labelenumii}{\theenumii}

\begin{document}

\maketitle

\section*{Reviewer 1}\label{sec:reviewer-1}

\begin{enumerate}
\item\textbf{Comment:} \label{comment:shortersec2}
	This paper is concerned with carrier phase and amplitude estimation for PSK modulated signalling.
The considered least squares estimator seems to be only a minor extension of the one from [7]. Yet, more than 4 pages are dedicated to its derivation. Perhaps this section could be made more concise.
	\\\textbf{Response:}
%\comment{look at shortening sec 2}

We have left the description of the algorithm intact.  The primary reason for this is that much of the notation used during the analysis of the estimator is most naturally introduced during the description of the algorithm.  To qualify this for the reader we have included the following sentences at the beginning of Section~\zref{sec:least-squar-estim}.

``In this section we derive Mackenthun's algorithm to compute the least squares estimators of the carrier phase and amplitude~\cite{Mackenthun1994}.  Mackenthun specifically considered the non data-aided setting, so we modify the algorithm to include pilot symbols.  Although the modification is straightforward, much of the notation that will be used for analysing the estimator in Section~\ref{sec:stat-prop-least} is most naturally introduced during the description of the algorithm.  For this reason, and so that our paper is self contained, we describe the algorithm in full.''

@@

	\\\textbf{Response:}
	We thank the reviewer for pointing out the missing definition of $f(r, \phi)$ and have added it. %Please note that in response to comment \ref{R4it2e} of reviewer 4, Sec.\ III has been slightly shortened and has been moved to the appendix.
\item\textbf{Comment:} 
	In the introductory section (paragraph before Eq.\ 3), the authors are talking about ``noncoherent detection where no pilot symbols exist''. In my mind this is somewhat confusing, because parameter estimation without knowing the underlying data is usually denoted as non-data aided (NDA) or decision-directed (DD), whereas when these data are known we have a data-aided (DA) situation\cite{Mengali_andre_synchro_book}. On the other hand, coherent/noncoherent scenarios are given if the carrier phase is known/unknown to the receiver station.

@@

\begin{enumerate}
\item\textbf{Comment:} 
	Figs.1 to 3 illustrate the evolution of the phase MSE vs.\ SNR. Only in case that $|P| = 0$,
the LS results are compared to an alternative solution represented by the Viterbi and Viterbi (VV) algorithm. Why not doing the same with $|P| > 0$ by the following approach, which is definitely much simpler from the computational point of view (complexity in the order $O(L)$ instead of $O(L \log(L))$ required by LS) and the jitter performance will probably differ not very much from that of the LS variant (in Figs.~1 to 3 this can be seen for $|P| = 0$):

@@

	\item
		If SNR $<$ SNRth, then the DA algorithm has to be used, else the VV algorithm must be applied.
	\item
		The SNRth is a threshold value known in advance, e.g., by the intersection of the Cramer-Rao lower bounds (CRLBs) for DA and NDA phase estimation\cite{Mengali_andre_synchro_book,950345}.

@@

	\item
		The SNR estimates might be obtained by a simple method based on second- and fourth-order moments~\cite{Pauluzzi2000}.

@@

		The same philosophy applies also to the estimation of amplitudes; for carrier- blind NDA results, the moment-based approach used for SNR estimation can be exploited.
	\end{itemize}
	\textbf{Response:}
The non data-aided estimation of SNR is a significant difficulty when implementing this approach.  As observed by simulation in~\cite{Pauluzzi2000} known non data-aided estimators of SNR appear to be biased.  The bias is often small when the noise variance is small, but typically becomes significant as the noise variance increases.  This makes it difficult to estimate SNR so as to know when to use a DA algorithm with only the available pilots and when to use the VV algorithm.  Analysis of the bias of the chosen non data-aided SNR estimator is required in order to justify this approach.  To date the bias and variance of non data-aided estimators of SNR (or equivalently carrier amplitude) appear to have been measured only by Monte-Carlo simulation~\cite{Pauluzzi2000}.  In the abstract of our new paper we point out that our of paper

``
appears to be the first time that the statistical properties of a non data-aided estimator for carrier amplitude have been analysed theoretically.
''

and in the new introduction we affirm that

``
while the specific results given here apply only to the least square estimator, the techniques that we develop to derive these results are now able to be deployed for the analysis of other estimators.
''

Thus, the techniques in our paper can potentially be deployed to accurately model the bias and variance of a non data-aided SNR estimator in order to implement the reviewers proposal.  We confirm however, that analysing and implementing the reviewers suggested technique is well beyond the scope of the present paper which places primary interest in analysing the least squares estimator.

\item\textbf{Comment:} 
	By detailed inspection of Fig.\ 4 showing the jitter variance for amplitude estimates, it is easily observed that $|P| < L$ performs better than $|P| = L$ in the medium-to-low SNR range. In my mind this is somewhat strange, since it should be the other way round (similar to the evolution of the phase curves in Figs.\ 1Ð3 and 5). Can you comment on this phenomenon ?
	\\\textbf{Response:}
In the original manuscript we plotted the \emph{variance} of the amplitude estimator, not the mean square error. That is, the solid lines where the theoretical variance of $\hat{m}_L = \hat{\rho} - \rho_0 G(0)$ and each point from the Monte-Carlo simulations was computed as the unbiased error $\tfrac{1}{T}\sum_{i=1}^T\big(\hat{\rho}_i - \rho_0G(0)\big)^2$.  We chose this approach in order to highlight the correctness of our asymptotic theory. This was described in the 4th paragraph of the old Section~8 containing the simulations section.

We do agree with the reviewer this this plot by itself initially appears misleading.  To alleviate in the new manuscript we instead plot the average square error computed in the usual way by Monte-Carlo simulation and the solid line is given by the bias plus variance squared.  The following paragraph has also be placed in the simulations section (Section~\zref{sec:simulations}) in the new manuscript.

@@

``Figure~\zref{fig:plotampBPSK} %,~\ref{fig:plotampQPSK}~and~\ref{fig:plotamp8PSK} 
shows the mean square error of the amplitude estimator when $M=2$ and $L=2048$ and when the number of pilots symbols is $\abs{P} = 0, \tfrac{L}{2}, L$.  The solid lines show the mean square error predicted by the bias $\rho_0\big(1 - G(0)\big)$ squared plus the variance of $\hat{m}_L = \hat{\rho} - \rho_0G(0)$ predicted by Theorem~\zref{thm:normality}.  That is, the solid lines are given by
\[
\rho_0^2\big(1 - G(0)\big)^2 + \rho_0^2\frac{p B_1 + d B_2}{L}. 
\]
When SNR is large $G(0)$ is close to $1$ and the bias of the amplitude estimator is small.  However, $G(0)$ grows without bound as the variance of the noise increases, so the bias is significant when SNR is small.  Figure~\zref{fig:plotampBPSK} suggests our asymptotic theory to be an excellent predictor of the mean square error of the amplitude estimator.
''

\item\textbf{Comment:} 
	In a much more powerful approach to the topic addressed in the current paper, the pilot sequence would be employed for a coarse estimation/correction of the carrier phase of the payload data. In the sequel, by embedding carrier phase recovery into an iterative (turbo) decoding scheme\cite{Lottici2004,Noels2005} or any other error correcting method~\cite{580211,974597}, the jitter performance is close to the CRLB for DA estimation even at very small SNRs, which are mainly determined by the convergence region of the (turbo) decoder. By taking into account the fact that modern and powerful communication systems are not conceivable without any FEC scheme, it is not clear to me why I should choose your algorithm in such a context.
	\\\textbf{Response:}
	We have added the following paragraph to the introduction describing the need for sufficiently accurate coarse estimates to initialise an iterative code-aided synchronisation stage. %citing [d, e].

``In the last decade, iterative methods for channel estimation that exploit the error correcting code used by the transmitter have appeared~\cite{Lottici2004,Noels2005,Herzet_turbo_synch_proc_IEEE_2007,Herzet_timingturbosynch_2007,Herzet_framework_turbo_sync_2007}.  These methods often go by the name of~\emph{turbo sychronisation}~\cite{Herzet_turbo_synch_proc_IEEE_2007}.  
These estimators typically apply the expectation maximisation (EM) algorithm under a Gaussian assumption regarding the noise sequence $\{w_i\}$.  A key problem is that the EM algorithm converges correctly only if initialised at some channel estimate sufficiently close to the true channel parameters.  Methods for efficiently obtaining a close initial channel estimate are still required and because of this the analysis of such methods remains of interest.  For the purpose of carrier phase and amplitude estimation, the least squares estimator we describe here is a good candidate as an initialiser for a turbo synchronisation procedure.''

\comment{BLERG, add LDPC turbo sync sims}

@@

\begin{enumerate}
\item
	\textbf{Comment:} 
The paper studies the estimation of amplitude and phase due to the distortion introduced by the channel in communication systems with $M$-PSK signals. The estimator is derived from the observation of $L$ noisy samples taken at a rate of one sample per symbol. Part of the $L$ observed samples corresponds to pilot transmitted symbols, that are already known a priori by the receiver, while the rest of samples are associated with data symbols which are unknown at the receiver. The focus of the article is to determine the amplitude and phase without considering the estimation of the transmitted data like this problem is usually treated in the literature. The work modifies the Mackenthun algorithm to include samples corresponding to pilot symbols while the Mackenthun paper calculates a least squares estimator for a similar problem but without pilot symbols.

The major contribution of this paper is the description of the asymptotic properties of the least squares estimator for the phase and amplitude. Although it seems technically correct, the authors do not make clear the relevance of their work to signal processing applications and to timely problems. They limit themselves in presenting several lemmas and proofs. Therefore, I cannot recommend this manuscript for publication in the IEEE Transactions on Signal Processing.
	\\\textbf{Response:}
We thank their reviewer for the comments. We have greatly expanded the introduction of the manuscript to make the practical and theoretical importance of our of results clear.
\end{enumerate}

\section*{Reviewer 4}

\begin{enumerate}
\item\textbf{Comment:}
	This work addresses the problem of carrier phase and amplitude estimation for $M$-ary PSK signals. Based on a modification of the Mackhentun's algorithm by inserting pilots, the main contribution of this manuscript is in fact the study of the asymptotic behavior of the least squares estimator for carrier phase and amplitude. Although the derivations seems correct, it is not clear to the reader why this method should be applicable in a communications context.  A cconsiderable part of the manuscript is dedicated to proofs of Lemmas and Theorems, which makes the paper rather difficult to read. There are several methods in the litterature that proposes the use of pilots to estimate the carrier phase followed by efficient decoding such as turbo decoding. How does the proposed method compare with the existing uncoherent methods that do not require estimation of the amplitudes? Even if the data are coded, what are the gains in terms of data recovery of the proposed method and traditional solutions? In my opinion, the authors fail to convincing on this aspect and this is an important point.
	\\\textbf{Response:}
We thank their reviewer for their comments. We have greatly expanded the introduction of the manuscript to make the practical and theoretical importance of our of results clear.

%	\comment{Stress practical relevance by showing ASRP PER results. In particular ST1, where the estimators are used to seed turbo sync. Mention trials and patents\dots}

@@

%	\comment{Not clear what the reviewer means by ``existing uncoherent methods that do not require estimation of the amplitudes''.}

\item
Some additional comments:
\begin{enumerate}
\item\textbf{Comment:} 
	The authors should better motivate the interest in using their proposed method in actual communication systems. The lack of motivation for addressing the problem raised in the manuscript is a weak aspect of this work.
	\\\textbf{Response:}
We have expanded the introduction of the paper to better motivate the work.  In particular, the following paragraph has been added:

``That this paper contains novel theoretical developments does not imply that our results are of no immediate practical interest.  The design of our estimator was motivated by a project at the University of South Australia's Institute for Telecommunications Research to develop a narrowband multiuser communication system to collect sensor network data using a satellite in orbit.  This estimator has been implemented on a prototype that was successfully trialled from April to July of 2013 with signals from multiple low power ground-based transmitters being successfully received and decoded by both aircraft and satellites in orbit~\cite{ASRPpromovideo}.  Our theoretical analysis aided design of this system.  Based on our Theorem~\zref{thm:normality} in Section~\zref{sec:stat-prop-least} the performance of the least squares estimator can be accurately predicted without the need for time consuming Monte-Carlo simulations.  This provides easy to use and rigourously justified methods for choosing systems parameters, such as the number and placement of pilot symbols, and also provides useful information to other reciever components, such as channel tracking systems.''

\item\textbf{Comment:} 
	The simulation results should bring comparisons with existing methods that use error correction and turbo decoding techniques, which have shown to perform reasonably well under adverse situations and low SNR levels.
	\\\textbf{Response:}
We have added the following paragraph to the introduction

\comment{Run turbo sync simulations and write something on this}

%It can be reasonably expected that the initialiser that is a more accurate estimator of carrier phase and amplitude with also be a better initialiser for a turbo synchronisation proceedure and will thus have a smaller bit error rate. So, for example, we expect our least square estimator to be a beter initialiser at moderate and high SNR and a data-aided approach to be better at low SNR provided there are sufficiently many pilots symbols.

% For this reason that we do not expect simulations of this sort to yield substantially more insight than the mean square error plots that we have presented in the paper.  The primary purpose of our paper is the statistical analysis of the least squares estimator and we feel that

%	\comment{Show PER ASRP ST1 turbo sync results. Comparison with other estimators?}
\item\textbf{Comment:} 
	In Fig.\ 4, the influence of $P$ on the performance is not clear to me and does not seem to be in accordance with other results in the manuscript.
	\\\textbf{Response:} 
	Please note that Fig.\ 4 in the original submission showed the variance of the amplitude estimator $\hat\rho$ (after bias removal), rather than the estimator MSE. This was done to validate the theoretical value of the asymptotic variance of $\hat\rho$ predicted by Theorem~2 against Monte-Carlo simulations. For consistency with the other figures and for easier interpretation, we have replaced the variance plot by a plot of the MSE of $\hat\rho$.
\item\textbf{Comment:} 
	It is important that the authors address an appropriate method for correcting the bias, as they mentioned in the text. A solution to provide the receiver with an estimate of $G(0)$ must be addressed in order to make this contribution applicable in a real communications context.
	\\\textbf{Response:} 
	Biased estimators are quite common in the communications literature and in fact, are often used in practice. Prime examples are state-of-the-art signal-to-noise ratio (SNR) estimators for $M$-PSK modulated signals such as those discussed by Pauluzzi and Beaulieu \cite{Pauluzzi2000}. With the exception of the data-aided SNR estimators, all of the estimators in \cite{Pauluzzi2000} are biased and particularly at low SNR, the bias can be large. To the best of the authors' knowledge, the value of the bias of these estimators has only been determined by computer simulation and has not been analysed theoretically. In contrast, the asymptotic analysis of the amplitude estimator $\hat\rho$ presented in Theorem~1 of our manuscript results in a closed-form expression for the bias and thus, lays the foundation for improving the estimator by addressing the bias. However, removal of the bias is beyond the scope of this manuscript.
\item\textbf{Comment:} \label{R4it2e}
	Some sectioms can be shortened to make the work more readable and concise. This is the case of Section II for instance. Section III is not necessary. It is too short and the main concepts are well known. A reference to a textbook would be sufficient. Section VII is also too short and is not justified.
	\\\textbf{Response:}

The original Sec.~VII describing the Gaussian noise case has been removed from the revised manuscript.

We agree with the reviewer that circularly symmetric random variables are well known, but would like to point out that section specifically discussed random variables of the form $Re^{j\Phi}=1{+}W$, where $W$ is circularly symmetric. The joint pdf of $R$ and $\Phi$ is a crucial ingredient for the asymptotic analysis presented in the manuscript and therefore, we feel that this material must be retained. However, we have modified the way that this material is introduced.  It is now introduced as needed when the asymptotic properties of the least squares estimator are derived.

Regarding Section~II we refer the reader to our response to comment~\eqref{comment:shortersec2} of reviewer~\ref{sec:reviewer-1}.

% Section~II describes our modification of Mackethun's algorithm to compute the least squares estimator.  Although the modification is simple we have left this section intact since much of the notation required for the statistical analysis of the least squares estimator is most naturally introduced during the description of the algorithm.  To qualify this for the reader we have included the following sentence at the start of Section~\zref{sec:least-squar-estim}.

% ``In this section we derive Mackenthun's algorithm to compute the least squares estimators of the carrier phase and amplitude~\cite{Mackenthun1994}.  Mackenthun specifically considered the non data-aided setting, so we modify the algorithm to include pilot symbols.  Although the modification is straightforward, much of the notation that will be used for analysing the estimator in Section~\zref{sec:stat-prop-least} is most naturally introduced during the description of the algorithm.  For this reason, and so that our paper is self contained, we describe the algorithm in full.''

%For example, equations~\eqref{eq:SSdatasymbols},~\eqref{eq:SSallparams} and~\eqref{eq:hatdfinxtheta} are all used during the

\end{enumerate}
\end{enumerate}

### 2013-11-19 paper.tex
more additions to response/paper

More ambitiously, our techniques might allow analysis of non data-aided channel estimators for modern communications systems such as mulitple anntannae systems, systems employing orthogonal frequency division multiplexing, and systems that must widthstand rapidly time varying channels~\cite{4138046,Heath99exploitinginput,inthft:204506}.  We see the results presented here as a starting point for a potentially potent approach to the analysis of channel estimators that is complementary to the commonplace method of comparing results of Monte-Carlo simulations with Cramer-Rao lower bounds.

@@

In the last decade, iterative methods for channel estimation that exploit the error correcting code used by the transmitter have appeared~\cite{Lottici2004,Noels2005,Herzet_turbo_synch_proc_IEEE_2007,Herzet_timingturbosynch_2007,Herzet_framework_turbo_sync_2007}.  These methods often go by the name of~\emph{turbo sychronisation}~\cite{Herzet_turbo_synch_proc_IEEE_2007}.  
These estimators typically apply the expectation maximisation (EM) algorithm under a Gaussian assumption regarding the noise sequence $\{w_i\}$.  A key problem is that the EM algorithm converges correctly only if initialised at some channel estimate sufficiently close to the true channel parameters.  Methods for efficiently obtaining a close initial channel estimate are still required and because of this the analysis of such methods remains of interest.  For the purpose of carrier phase and amplitude estimation, the least squares estimator we describe here is a good candidate as an initialiser for a turbo synchronisation proceedure.  %However, in some cases, it may be that our estimator is accurate enough, and that the gain obtained by applying the EM algorithm will not justify the extra computational cost.

@@

The paper is organised in the following way.  Section~\ref{sec:least-squar-estim} extends Mackenthun's algorithm for the case when both pilot symbols and data symbols exist.  Section~\ref{sec:circ-symm-compl} describes properties of complex random variables that we need.  Section~\ref{sec:stat-prop-least} states two theorems that describe the statistical properties of the least squares estimator of carrier phase and amplitude.  %We show, under reasonably general assumptions about the distribution of the noise $w_1,\dots,w_L$, that $\langle\theta_0 - \hat{\theta}\rangle_\pi$ converges almost surely to zero and that $\sqrt{L}\langle\theta_0 - \hat{\theta}\rangle_\pi$ is asymptotically normally distributed as $L\rightarrow \infty$.  However, $\hat{\rho}$ is not a consistent estimator of the amplitude $\rho_0$.  The asymptotic bias of $\hat{\rho}$ is small when the signal to noise ratio (SNR) is large, but the asymptotic bias is significant when the SNR is small.  
Sections~\ref{sec:proof-almost-sure} and~\ref{sec:proof-asympt-norm} provide proofs of the theorems stated in Section~\ref{sec:stat-prop-least}.  %In Section~\ref{sec:gaussian-noise-case} we consider the special case when the noise is Gaussian.  In this case, our expressions for the asymptotic distribution can be simplified.  Section~\ref{sec:simulations} presents the results of Monte-Carlo simulations.  
These simulations agree with the derived asymptotic properties.

@@

In this section we derive Mackenthun's algorithm to compute the least squares estimators of the carrier phase and amplitude~\cite{Mackenthun1994}.  Mackenthun specifically considered the non data-aided setting, so we modify the algorithm to include pilot symbols.  Although the modification is straightforward, much of the notation that we will use for analysing the estimator in Section~\ref{sec:stat-prop-least} is naturally introduced during the description of the algorithm.  For this reason, and so that our paper is self contained, we describe the algorithm in full.  For the purpose of analysing computational complexity, we will assume that the number of data symbols $\abs{D}$ is proportional to the total number of symbols $L$, so that, for example, $O(L) = O(\abs{D})$.  In this case Mackentun's algorithm requires $O(L \log L)$ arithmetic operations.  This complexity arises from the need to sort a list of $\abs{D}$ elements.  
%We use order notation in the standard way, that is, for functions $h$ and $g$, we write $h(N) = O(g(N))$ to mean that there exists a constant $K > 0$ and a finite $N_0$ such that $h(N) \leq K g(N)$ for all $N > N_0$.

@@

Let,
\begin{equation}\label{eq:SSfk}
SS(f_k) = A - \frac{1}{L}\abs{Y_k}^2,
\end{equation}
where, 
\begin{align*}
Y_k = Y( f_k ) &= \sum_{i \in P} y_i p_i^*  + \sum_{i \in D} y_i f_{ki}^* \\
&= B + \sum_{i \in D}g_{ki},
\end{align*}
where $B = \sum_{i \in P} y_i p_i^*$ is independent of the data symbols, and $f_{ki}$ denotes the $i$th symbol in $f_k$, and for convenience, we put $g_{ki}  = y_i f_{ki}^*$.  Letting $g_{k}$ be the sequence $\{g_{ik}, i \in D\}$ we have, from~\eqref{eq:fkrec}, that $g_k$ satisfies the recursive equation
\[
g_{k+1} = g_{k} e_{\sigma(k)}^*,
\]
where $g_{k} e_{\sigma(k)}^*$ indicates the sequence $g_k$ with the $\sigma(k)$th element replaced by $g_{k \sigma(k)}e^{j2\pi/M}$.  Now,
\[
Y_0 = B + \sum_{i \in D} g_{0i}
\] 
can be computed in $O(L)$ operations, and
\begin{align*}
Y_1 &= B + \sum_{i \in D} g_{1i} \\
&= B +  (e^{j2\pi/M} - 1)g_{0\sigma(0)} + \sum_{i \in D} g_{0i} \\
&= Y_0 + \eta g_{0\sigma(0)},
\end{align*}
where $\eta = e^{j2\pi/M} - 1$.  In general,
\[
Y_{k+1} = Y_k + \eta g_{k\sigma(k)}.
\]
So, each $Y_k$ can be computed from its predecessor $Y_{k-1}$ in $O(1)$ arithmetic operations.  Given $Y_k$, the value of $SS(f_k)$ can be computed in $O(1)$ operations using~\eqref{eq:SSfk}.  Let $\hat{k} = \arg\min SS(f_k)$.  The least squares estimator of $a_0$ is then computed according to~\eqref{eq:hata},
\begin{equation}\label{eq:ahatYhat}
\hat{a} = \frac{1}{L} Y_{\hat{k}}.
\end{equation}
Pseudocode is given in Algorithm~\ref{alg:loglinear}.  Line~\ref{alg_sortindices} contains the function $\operatorname{sortindices}$ that, given $z = \{z_i, i \in D\}$, returns the permutation $\sigma$ as described in~\eqref{eq:sigmasortind}.  The $\operatorname{sortindices}$ function requires sorting $\abs{D}$ elements.  This requires $O(L \log L)$ operations.  The $\operatorname{sortindices}$ function is the primary bottleneck in this algorithm when $L$ is large.  The loops on lines~\ref{alg_loop_setup} and~\ref{alg_loop_search} and the operations on lines~\ref{alg_Y} to lines~\ref{alg_Q} all require $O(L)$ or less operations.  %In the next sections we will show how to the sortinces function can be avoided.  This leads to an algorithm that requires only $O(L)$ operations.

@@

\section{Circularly symmetric complex random variables}\label{sec:circ-symm-compl}

Before describing the statistical properties of the least squares estimator, we first require some properties of complex valued random variables.  A complex random variable $W$ is said to be \emph{circularly symmetric} if its phase $\angle{W}$ is independent of its magnitude $\abs{W}$ and if the distribution of $\angle{W}$ is uniform on $[0,2\pi)$.  That is, if $Z \geq 0$ and $\Theta \in [0,2\pi)$ are real random variables such that $Ze^{j\Theta} = W$, then $\Theta$ is uniformly distributed on $[0,2\pi)$ and is independent of $Z$.  If the probability density function (pdf) of $Z$ is $f_Z(z)$, then the joint pdf of $\Theta$ and $Z$ is 
\[
f_{Z,\Theta}(z,\theta) = \frac{1}{2\pi}f_Z(z).
\]
% Observe that for any real number $\phi$, the pdf of $W$ and $e^{j\phi}W$ are the same, that is, the pdf is invariant to phase rotation.  If $\expect\abs{W} = \expect Z$ is finite, then $W$ has zero mean because
% \begin{align*}
%  \expect W &= \int_{0}^{2\pi} \int_{0}^\infty z e^{j\theta} f_{Z,\Theta}(z,\theta) dz d\theta \\
%  &= \frac{1}{2\pi} \int_{0}^{2\pi} e^{j\theta} \int_{0}^\infty z f_Z(z) dz d\theta \\
%  &= \frac{1}{2\pi}\expect Z \int_{0}^{2\pi} e^{j\theta} d\theta = 0.
%  \end{align*}
% If $X$ and $Y$ are real random variables equal to the real and imaginary parts of $W = X + jY$ then the joint pdf of $X$ and $Y$ is
% \[
% f_{X,Y}(x,y) = \frac{f_Z(\sqrt{x^2 + y^2})}{2\pi \sqrt{x^2 + y^2}}.
% \]
%from which follows
%\begin{equation}\label{eq:fzandfZYwithz}
%f_Z(z) = 2\pi z f_{X,Y}(z,0) = 2\pi z f_{X,Y}(0,z)
%\end{equation}
%for all $z > 0$.

>>>>

More ambitiously, our techniques might allow analysis of non data-aided channel estimators for modern communications systems such as mulitple anntannae systems, systems employing orthogonal frequency division multiplexing, and systems that must widthstand rapidly time varying channels~\cite{4138046,inthft:204506}.  We see the results presented here as a starting point for a potentially potent approach to the analysis of channel estimators that is complementary to the commonplace method of comparing results of Monte-Carlo simulations with Cramer-Rao lower bounds.

@@

In the last decade, iterative methods for channel estimation that exploit the error correcting code used by the transmitter have appeared~\cite{Lottici2004,Noels2005,Herzet_turbo_synch_proc_IEEE_2007,Herzet_framework_turbo_sync_2007}.  These methods often go by the name of~\emph{turbo sychronisation}~\cite{Herzet_turbo_synch_proc_IEEE_2007}.  
These estimators typically apply the expectation maximisation (EM) algorithm under a Gaussian assumption regarding the noise sequence $\{w_i\}$.  A key problem is that the EM algorithm converges correctly only if initialised at some channel estimate sufficiently close to the true channel parameters.  Methods for efficiently obtaining a close initial channel estimate are still required and because of this the analysis of such methods remains of interest.  For the purpose of carrier phase and amplitude estimation, the least squares estimator we describe here is a good candidate as an initialiser for a turbo synchronisation proceedure.  %However, in some cases, it may be that our estimator is accurate enough, and that the gain obtained by applying the EM algorithm will not justify the extra computational cost.

@@

The paper is organised in the following way.  Section~\ref{sec:least-squar-estim} extends Mackenthun's algorithm for the case when both pilot symbols and data symbols exist.  %Section~\ref{sec:circ-symm-compl} describes properties of complex random variables that we need.  
Section~\ref{sec:stat-prop-least} states two theorems that describe the statistical properties of the least squares estimator of carrier phase and amplitude.  %We show, under reasonably general assumptions about the distribution of the noise $w_1,\dots,w_L$, that $\langle\theta_0 - \hat{\theta}\rangle_\pi$ converges almost surely to zero and that $\sqrt{L}\langle\theta_0 - \hat{\theta}\rangle_\pi$ is asymptotically normally distributed as $L\rightarrow \infty$.  However, $\hat{\rho}$ is not a consistent estimator of the amplitude $\rho_0$.  The asymptotic bias of $\hat{\rho}$ is small when the signal to noise ratio (SNR) is large, but the asymptotic bias is significant when the SNR is small.  
Sections~\ref{sec:proof-almost-sure} and~\ref{sec:proof-asympt-norm} provide proofs of the theorems stated in Section~\ref{sec:stat-prop-least}.  %In Section~\ref{sec:gaussian-noise-case} we consider the special case when the noise is Gaussian.  In this case, our expressions for the asymptotic distribution can be simplified.  Section~\ref{sec:simulations} presents the results of Monte-Carlo simulations.  
These simulations agree with the derived asymptotic properties.

@@

In this section we derive Mackenthun's algorithm to compute the least squares estimators of the carrier phase and amplitude~\cite{Mackenthun1994}.  Mackenthun specifically considered the non data-aided setting, so we modify the algorithm to include pilot symbols.  Although the modification is straightforward, much of the notation that will be used for analysing the estimator in Section~\ref{sec:stat-prop-least} is most naturally introduced during the description of the algorithm.  For this reason, and so that our paper is self contained, we describe the algorithm in full.  For the purpose of analysing computational complexity, we will assume that the number of data symbols $\abs{D}$ is proportional to the total number of symbols $L$, so that, for example, $O(L) = O(\abs{D})$.  In this case Mackentun's algorithm requires $O(L \log L)$ arithmetic operations.  This complexity arises from the need to sort a list of $\abs{D}$ elements.  
%We use order notation in the standard way, that is, for functions $h$ and $g$, we write $h(N) = O(g(N))$ to mean that there exists a constant $K > 0$ and a finite $N_0$ such that $h(N) \leq K g(N)$ for all $N > N_0$.

@@

Let,
\begin{equation}\label{eq:SSfk}
SS(f_k) = A - \frac{1}{L}\abs{Y_k}^2,
\end{equation}
where, 
\[
Y_k = Y( f_k ) = \sum_{i \in P} y_i p_i^*  + \sum_{i \in D} y_i f_{ki}^* = B + \sum_{i \in D}g_{ki},
\]
where $B = \sum_{i \in P} y_i p_i^*$ is independent of the data symbols, and $f_{ki}$ denotes the $i$th symbol in $f_k$, and for convenience, we put $g_{ki}  = y_i f_{ki}^*$.  Letting $g_{k}$ be the sequence $\{g_{ik}, i \in D\}$ we have, from~\eqref{eq:fkrec}, that $g_k$ satisfies the recursive equation
\[
g_{k+1} = g_{k} e_{\sigma(k)}^*,
\]
where $g_{k} e_{\sigma(k)}^*$ indicates the sequence $g_k$ with the $\sigma(k)$th element replaced by $g_{k \sigma(k)}e^{j2\pi/M}$.  Now,
\[
Y_0 = B + \sum_{i \in D} g_{0i}
\] 
can be computed in $O(L)$ operations, and
\begin{align*}
Y_1 &= B + \sum_{i \in D} g_{1i} \\
&= B +  (e^{j2\pi/M} - 1)g_{0\sigma(0)} + \sum_{i \in D} g_{0i} \\
&= Y_0 + \eta g_{0\sigma(0)},
\end{align*}
where $\eta = e^{j2\pi/M} - 1$.  In general,
\[
Y_{k+1} = Y_k + \eta g_{k\sigma(k)}.
\]
So, each $Y_k$ can be computed from its predecessor $Y_{k-1}$ in $O(1)$ arithmetic operations.  Given $Y_k$, the value of $SS(f_k)$ can be computed in $O(1)$ operations using~\eqref{eq:SSfk}.  Let $\hat{k} = \arg\min SS(f_k)$.  The least squares estimator of $a_0$ is then computed according to~\eqref{eq:hata},
\[
\hat{a} = \frac{1}{L} Y_{\hat{k}}.
\]
Pseudocode is given in Algorithm~\ref{alg:loglinear}.  Line~\ref{alg_sortindices} contains the function $\operatorname{sortindices}$ that, given $z = \{z_i, i \in D\}$, returns the permutation $\sigma$ as described in~\eqref{eq:sigmasortind}.  The $\operatorname{sortindices}$ function requires sorting $\abs{D}$ elements.  This requires $O(L \log L)$ operations.  The $\operatorname{sortindices}$ function is the primary bottleneck in this algorithm when $L$ is large.  The loops on lines~\ref{alg_loop_setup} and~\ref{alg_loop_search} and the operations on lines~\ref{alg_Y} to lines~\ref{alg_Q} all require $O(L)$ or less operations.  %In the next sections we will show how to the sortinces function can be avoided.  This leads to an algorithm that requires only $O(L)$ operations.

@@

In this section we describe the asymptotic properties of the least squares estimator.  We first require some properties of complex valued random variables.  A complex random variable $W$ is said to be \emph{circularly symmetric} if its phase $\angle{W}$ is independent of its magnitude $\abs{W}$ and if the distribution of $\angle{W}$ is uniform on $[0,2\pi)$.  That is, if $Z \geq 0$ and $\Theta \in [0,2\pi)$ are real random variables such that $Ze^{j\Theta} = W$, then $\Theta$ is uniformly distributed on $[0,2\pi)$ and is independent of $Z$.  If the probability density function (pdf) of $Z$ is $f_Z(z)$, then the joint pdf of $\Theta$ and $Z$ is 
\[
f_{Z,\Theta}(z,\theta) = \frac{1}{2\pi}f_Z(z).
\]
% Observe that for any real number $\phi$, the pdf of $W$ and $e^{j\phi}W$ are the same, that is, the pdf is invariant to phase rotation.  If $\expect\abs{W} = \expect Z$ is finite, then $W$ has zero mean because
% \begin{align*}
%  \expect W &= \int_{0}^{2\pi} \int_{0}^\infty z e^{j\theta} f_{Z,\Theta}(z,\theta) dz d\theta \\
%  &= \frac{1}{2\pi} \int_{0}^{2\pi} e^{j\theta} \int_{0}^\infty z f_Z(z) dz d\theta \\
%  &= \frac{1}{2\pi}\expect Z \int_{0}^{2\pi} e^{j\theta} d\theta = 0.
%  \end{align*}
% If $X$ and $Y$ are real random variables equal to the real and imaginary parts of $W = X + jY$ then the joint pdf of $X$ and $Y$ is
% \[
% f_{X,Y}(x,y) = \frac{f_Z(\sqrt{x^2 + y^2})}{2\pi \sqrt{x^2 + y^2}}.
% \]
%from which follows
%\begin{equation}\label{eq:fzandfZYwithz}
%f_Z(z) = 2\pi z f_{X,Y}(z,0) = 2\pi z f_{X,Y}(0,z)
%\end{equation}
%for all $z > 0$.
